Supplementary Materials to ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic''

### 2022-02-09 write-up/letters/response-ref-1-v1.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}
\usepackage{hyperref}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

We appreciate all of the very helpful comments from the editor, associate editor, and the two referees which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a statement from the referee report.

{\bf Response to Editor and Associate Editor}
\begin{enumerate}
\item {\it The referees have many specific concerns they are asking to address, and their expertise and viewpoints are important to revise this manuscript to improve its clarity and impact.}

\vspace{5mm}
We thank the editor-in-chief as this is an excellent point.  We now acknowledge that both topics are well known and cite relevant literature in Section 1.1; however, while infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.   By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it Some of these deal with the fast-evolving nature of the pandemic with new information emerging every week and changing variants with different characteristics also changing our understanding of the disease course and spread and timing, that affect the content in the paper. I understand that it is not feasible for a peer-reviewed paper like this to include all the latest information in a fast-changing environment like this in which understanding can change quickly, but it is important in this revision to update the KEY points as much as possible based on updated knowledge, but otherwise revise the language to make it “forward compatible” in terms of relevance as things continue to change, e.g. by acknowledging some details of spread dynamics may change with future variants, but presenting the current work based on a particular SARS-CoV-2 variant as illustrative of the issues that arise with each new variant and other viruses, as well. That is, recognizing it may not be feasible to keep updating based on emerging knowledge from each new strain, but to lock down the discussion context but add enough explanation so the reader can see its relevance for new strains.}

\vspace{5mm}
x
\vspace{5mm}

\item {\it The AE also mentions a few comments as well:}
\begin{itemize}
	\item {\it Title: As noted by one of the reviewers, the “paradox” is never discussed/defined in the manuscript. Consider changing to “Statistical estimation of Coronavirus Case Counts: Measurement…” or some alternative.}
	\item {\it Organization: The figures and tables are not located anywhere near much of their introduction/discussion. For example, Table 1 was discussed on page 7 and is itself located on page 22. Consider the reorganization such that Figures and Tables are located near their discussion.}
	\item {\it Discussion on the Indiana Random Sample: On page 22, table 1, it appears that the random sample is not similar to the census estimates, potentially indicating non-response bias.}
	\item {\it Additionally, there may be significant potentially non-ignorable variables (e.g., political opinions and rural status) that drive non-response and are not presented. Is there any comment/discussion on the success of the randomization to be used as the “gold-standard” in this example? Much of the results appear to rely on the SRS and it ignores the potential for nonresponse bias.}
\end{itemize}

\vspace{5mm}
Thank you for raising this issue.  We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.  Given currently available public data, one cannot correct at the daily level for reporting delays.  Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.  While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
\vspace{5mm}
\end{enumerate}
\newpage
{\bf Response to Reviewer 1}
I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).

\begin{enumerate}
\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure. Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing. Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random. The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}

\vspace{5mm}
Section XX now discusses the purposes of testing.
\vspace{5mm}

\item {\bf Section 1.1}
\begin{itemize}
	\item {\it For the first goal of the paper, it might be helpful to link the biases with those outlined in a set of papers from the Harvard Center for Disease Dynamics: \url{https://doi.org/10.1007/s10654-021-00727-7} and DOI:\url{10.1093/aje/kwaa188}}
	\vspace{5mm}
	We now connect with XX
	\vspace{5mm}
	\item {\it For the second goal of the paper, it would be useful to discuss how the proposed method differs from the more traditional epidemiologic method of obtaining a validation sample. As I see it, the chief distinction is that here the “gold standard” is not a different test (although one could argue that the gold standard being used here is a screening test), but rather it is a less biased selection mechanism. Clarifying the links with validation studies could be helpful for epidemiologists \& statisticians alike, as the use of a validation study to correct for selection bias is a fairly novel suggestion.}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
	\item {\it Finally, consider linking the arguments you are making with recent work on ‘target validity’ (see: \url{https://doi.org/10.1093/aje/kwy228}). Literature on target validity has largely been focused on specifying the problem of tackling internal and external validity together, but the method you propose here is a concrete attempt to address both types of validity and extends this conversation nicely.}
	\vspace{5mm}
	Validation subsection
	\vspace{5mm}
\end{itemize}
\item {\bf Section 2.1}
\begin{itemize}
	\item {\it Please clarify in the first paragraph the distinction between incubation period (time to symptom onset) and latent period (time to infectiousness). The viral load will be detectable by at least the end of the latent period which for SARS-CoV-2 occurs before the end of the incubation period. However, individuals are unlikely to present for diagnostic testing until after the end of the incubation period.}
	\vspace{5mm}
	We now connect with XX
	\vspace{5mm}
	\item {\it Please also clarify the terminology ‘active infection’ which in some instances appears to refer to individuals in the symptomatic period of infection, and in other instances appears to refer to individuals in the infectious period of infection. We know that these are not identical in SARS-CoV-2 infection and that difference is a key reason why screening and diagnostic tests are not easily comparable.}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
	\item {\it Consider adding some additional information on screening tests, particular anything relevant to the random sample conducted in Indiana which is used as the validation set. Were the screening tests conducted using nasopharyngeal swabs (in many places screening tests use nasal swabs instead and often self-collected samples)? Were the molecular tests used to assay the samples the same for screening and diagnostic tests? Differences in sample collection and testing processes could undermine the utility of this random sample as a validation set.}
	\vspace{5mm}
	Validation subsection
	\vspace{5mm}
\end{itemize}
\item {\bf Section 2.1}
\begin{itemize}
	\item {\it This section could benefit from a few remarks on why it is important / useful to focus on publicly available testing data. There are certainly distinct statistical issues, if public data are aggregated, but are the selection bias and measurement error issues different between public and government datasets? If so, how do the solutions presented here apply to government datasets? Can similar or the same approaches be used by health departments to assess the pandemic and guide responses?}
	\vspace{5mm}
	Clarify that Indiana released a lot of aggregate information.  Two distinct goals: (1) what can we say with additional surveys but not additional testing, (2) guide health departments who have access to
	\vspace{5mm}
	\item {\it A timeline of restrictions and testing eligibility would be a useful addition to Section 2.2.1. Also, what there no time period where testing in Indiana was restricted on the basis of travel history or suspected exposure?}
	\vspace{5mm}
	Remark wihtin validation subsection to contrast with validation sample.
	\vspace{5mm}
\end{itemize}

\end{enumerate}
\newpage

{\bf Response to AE3}
\begin{itemize}
\item {\it XX}

\vspace{5mm}
WXX
\vspace{5mm}

\end{itemize}

\end{letter}

\end{document}

### 2022-02-09 write-up/letters/round1/response-ref-1-v1.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right
\usepackage{amsthm, amsfonts}

\def\Y{{\bf Y}}

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{Professor
	Peter Glynn\\
	Editor, {\em Journal of Applied Probability}}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%\begin{center}
%\large
%\end{center}
%\vfill

\signature{Walter Dempsey\\
University of Michigan\\
Department of Biostatistics\\
1415 Washington Heights\\
Ann Arbor, MI 48103} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\vspace{5mm}

\newpage

{\bf Response to Referees}

We appreciate all of the very helpful comments from the editor-in-chief, four associate editors, and the referee which has helped improve the paper substantially. We have addressed the concerns as best as possible in this revision. We provide a further point-by-point explanation below. Any comments in {\it italics} indicate a quote from the referee report.

{\bf Response to Editor-in-Chief}
\begin{enumerate}
\item {\it Selection bias and measurement error are certainly well known.
Certainly worth pointing out as both apply here.  But both should be acknowledged as being very well known.}

\vspace{5mm}
We thank the editor-in-chief as this is an excellent point.  We now acknowledge that both topics are well known and cite relevant literature in Section 1.1; however, while infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.   By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it My reaction after reading the introduction was: "OK, we know that.
But, what do you suggest we do?  Throw up our hands and go home?
Just tell people, `We can't help you.  The data won't help us to
model the epidemic or to forecast what will happen. Forget the data,
we'll just have to wait and see what happens.'" I don't think that's
the message you want to convey.  And we certainly do not want non-
statisticians (interested in Covid-19 data) to read your paper and
advise public health officials against collecting data altogether
because it has too many biases!  Rather, I think you want to alert
the reader that adjustments need to be made before modeling the
data, and, more importantly, we may need to develop new models
for this situation where most of the outcomes refer to suspected
individuals, not asymptomatic cases.}

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective and may have given the wrong impression to non-statisticians.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods that provide a route forward.  We then apply these methods to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\item {\it BTW, this surely is not a new problem - likely HIV testing is
generally done on only those individuals suspected to be in
high-risk groups.  Nor is the problem of reporting delays;
CDC routinely adjusts monthly counts of AIDS for reporting delays.
You may need to reference that literature.}

\vspace{5mm}
Thank you for raising this issue.  We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.  Given currently available public data, one cannot correct at the daily level for reporting delays.  Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.  While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
\vspace{5mm}

\item {\it Another point that may be worth mentioning is the potential
effect of length-biased sampling: the longer an individual
has been exposed, the more likely (s)he is to seek symptomatic
testing).  This effect hopefully will be absent in context of
random selection of individuals for asymptomatic testing.}

\vspace{5mm}
This is an excellent point and was the motivation for considering how to address length-biased sampling and other potential reasons that make individuals more likely to seek out testing.  Here, we address this by building selection propensities that may depend both on symptom status and the length of exposure as reported by the individual at the time of testing.   The lack of data to address length-biased sampling is pointed out in Section 6 and guides the suggested real-world improvements to data collection that will enable our proposed approach to combine survey statistical methods with infectious disease modeling to improve our understanding of active infection rates over time.
\vspace{5mm}

\item {\it Incidentally, I see the most frustrating part of the case count
reports as lack of two critical covariates on each case: (i) age;
(ii) presence or absence of relevant co-morbidity (auto-immune
condition, suppressed immune system, compromised respiratory or
lung condition).  I think the case count data, adjusted for
reporting delays and incorporating undertainty due to measurement
error as best we can, would be far more useful with these two
covariates.  And if we can get a handle on how to adjust for
selection bias, even better.}

\vspace{5mm}
We agree wholeheartedly with this critique.  To try and improve covariate reporting, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. We acknowledge the limitations of available demographic information and use the Facebook symptom survey to try and impute the remaining important covariates.  For a separate stratified random sample from Indiana that we were able to access, these critical covariates are available and we show how this data can be used to improve over unweighted approaches.  Overall, the proposed methodology uses these critical covariates to account for selection bias.  Model-based approaches are extended to handle this information and combined to build doubly robust estimation strategies as well.  These new methods, aimed directly at this frustrating issue, motivate our practical recommendations to improve policy making and increase accuracy in estimation using case count data.
\vspace{5mm}

\end{enumerate}

\noindent {\bf Editor-in-Chief: Specific comments}

\begin{itemize}
\item {\it Abstract, l.3: "Throughout this" this what? "this pandemic"?}

\vspace{5mm}
Yes, we meant pandemic.  This has now been made explicit.
\vspace{5mm}

\item {\it p2, p.5-6: "For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored." Drastic oversimplification, don't you think?}

\vspace{5mm}
We agree that this is a simplification.  However, we continue to start in this setting as a means to help readers understand the statistical decomposition in Section 3.1 as well as the pseudo-likelihood approach detailed in Section 4.  Both are extended to account for the dynamic nature of the outbreak in subsequent sections.
\vspace{5mm}

\item {\it p3, l.-3: Generally, $X_j = \mu_j + \epsilon_j$.Here, it seems as those you are saying: $P_j = 1$ if $\epsilon_j \neq 0$ and $P_j = 0$ if $\epsilon_j = 0$, is that right?}

\vspace{5mm}
We are focused exclusively on the binary outcome setting where $Y_j \in \{0,1\}$ as one is either COVID-19 positive~($Y_j = 1$) or negative~($Y_j = 0$).  Under measurement error, we observe $Y_j^\star \in \{0,1\}$ which is equal to $Y_j (1-P_j) + (1-Y_j) P_j$ where $P_j$ is the binary indicator of observing the incorrect value.
\vspace{5mm}

\item {\it p5, l.-2:  I must have missed the def'n of $D_M$ -- I looked back
three times and still cannot find it.}

\vspace{5mm}
The term $D_M$ is defined in equation (3.1) and represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence.  Figure 2 visualizes the adjustment as a contour plot to help the reader understand how the adjustment
\vspace{5mm}

\item {\it You'll also need to define SIR.  See AE2's note.}

\vspace{5mm}
We apologize for the oversight.  We thank you and AE2 for bringing this to our attention.  We now mathematically define the susceptible-infected-recovered (SIR) model in Section 3.2 as well as the susceptible-exposed-infected-recovered (SEIR) in Section 4.2.
\vspace{5mm}

\item {\it p6, Fig 1 has no legend.  It also looks like most of the action is in the region (Sampling Frequency 0.5 to 5), (Odds Ratio 1-50). I also think that a plot with contour levels would both avoid the need for color altogether and would be much easier to read.  Did you try to read the labels on the axes?  It's impossible unless you magnify the pdf 8-fold.}

\vspace{5mm}
We have fixed Figure 1 (now Figure 2) to be a contour plot with appropriately sized labels on the axes. We apologize for needing a magnifying glass to read it before.
\vspace{5mm}

\item {\it p10, Fig 2: you definitely do not need color here; different
line types suffice.  Likewise for Fig 3 (p14) and Fig 4 (p15).}

\vspace{5mm}
We have fixed all figures to be in grayscale and use different line types.
\vspace{5mm}

\item {\it p18: Please follow AOAS Guidelines for references; see the
web page from where you submitted your MS.}

\vspace{5mm}
The references now follow AOAS Guidelines.
\vspace{5mm}

\item {\it p20: Appendices can go into a Supplementary File (unless
you have a strong reason for including them here after the References).}

\vspace{5mm}
The appendices have been moved to a supplementary file as suggested.
\vspace{5mm}

\end{itemize}
\newpage

{\bf Response to AE1}
\begin{itemize}
\item {\it I was confused about what prevalence and case numbers the author is addressing. My understanding is that there are primary two assays, one for serology (evidence of ever having the virus) measured in the blood, and the other is presence of the active virus measured by swabbing the throat.  The characteristics of the assays, in terms of measurement error, are different.  In addition, there exist several different laboratory procedures for doing these two types of assays with varying accuracy.  The paper does not seem to address these basic issues.}

\vspace{5mm}
We thank the reviewer for points this out. Section 2.1 now discusses diagnostic testing.  Given the revision is focused mostly on molecular tests, we limit our discussion to these types of assays and their scientific use in managing the pandemic.  We discuss reasons behind false positives/negatives and cite the most recent systematic reviews of sensitivity and specificity for RT-PCR tests.
\vspace{5mm}

\item {\it My primary concern relates to the fundamental mission of AOAS. The manuscript has very limited data analysis, limited primarily to pp.14-15, in part because the data are very limited: it is difficult to obtain data about the various aspects of the testing for this virus. The data analysis that is presented on pp.14-15 is mostly about accounting for sensitivity and specificity of the serology assays, not the testing for presence of the Covid-19 virus at a particular time.  There is an assumption with the serology that anyone who has had Covid-19 at some time would have antibodies. These assumptions need to be explicitly made.  I am not sure if anyone knows whether this assumption is even correct or to what degree it might be correct. How does this fit into the measurement error that is discussed and illustrated in the data analysis?}

\vspace{5mm}
We agree that the original manuscript was too focused on the scientific communication perspective although it did contain original derivations related to extending the statistical error decomposition of Meng (2018) to account for measurement error.  The revised manuscript now fits well within the mission of AOAS by focusing on novel methods and clear application to a current and important scientific problem. In particular, we use the decomposition to motivate new methodology that leverages auxiliary information to address selection bias. In Section 2, we now introduce the data streams that illustrate this methodology. First, we worked with the state of Indiana to release daily COVID-19 testing, case-count, and deaths by demographic strata. Second, we were granted access to the Facebook symptom survey which provides daily measures of important time-varying covariates such as the fraction of individuals currently experiencing COVID-19 related symptoms and/or in recent contact with a COVID-19 positive individual.  The Facebook survey provides auxiliary information that is used to construct selection propensities, which are then incorporated into an inverse-probability weighting method (introduced in Section 4.1) to estimate time-varying active infection rates. Of course, many infectious disease experts prefer compartmental model-based approaches built from differential equations such as the susceptible-exposed-infected-recovered (SEIR) model.  In Section 4.2, we build a novel compartmental model to take advantage of strata-level COVID-19 death count data.  In Section 4.3, we show how to combine these models with inverse weighting to construct doubly-robust estimates of active infection rates.  We extend the decompositions to the weighting and doubly robust setting which then guide our discussion of practical implications.
\vspace{5mm}

\item {\it The beginning of the manuscript describes the very important issue of selection bias.  But the discussion is regrettably limited.  It is unfortunate that the manuscript only refers to the Covid-19 data (not the analysis) and offers insufficient attention to what I view as the most important issue.  The manuscript could have offered both greater depth and wider coverage of selection bias and other issues raised by the author.}

\vspace{5mm}
We now expand our coverage of selection bias issue by considering what it will take to overcome it with respect to COVID-19 case count data.  The auxiliary information helps address selection bias by allowing us to build selection propensity models while the doubly robust methodology helps us incorporate well known infectious disease models.  Overall, the paper has moved from just acknowledging the issue of selection bias and measurement error -- albeit from a new lens using the decomposition of Meng (2018)) -- to providing solutions and practical recommendations based on the proposed methodology.
\vspace{5mm}

\item {\it All in all, I don't see enough innovation in the data analysis and methods described here that would make AOAS a good home for it.  This manuscript appears to be a methodological review of data issues related to the measuring of the public health impact of the Covid-19 pandemic. I think this manuscript would be better suited for quantitative epidemiology journal or a quantitative journal in infectious diseases.}

\vspace{5mm}
As stated above, we agree that the original manuscript was too focused on the scientific communication perspective.  The revised manuscript, with its methodological innovation and application to a current and important relevant scientific application, now fits well within the mission of AOAS.
\vspace{5mm}

\end{itemize}

\newpage

{\bf Response to AE2}
\begin{enumerate}
\item {\it I read the paper (AOAS2006-048) but did not study any derivations. It seems straightforward, but I'm not well acquainted with the statistical area and do not have much intuition for the author's development. I'm not suggesting that it is incorrect. To get a reasonable understanding of it, I would have to work carefully through the logic and details.  In the interest of time, I've not done so.}

\vspace{5mm}
The original derivations were straightforward and followed from some arithmetic calculations, but the main point was the resulting statistical error decomposition.  While infectious disease researchers and survey statisticians have been aware of the perils of self-selection and measurement error, the interplay in this context has been less well studied.  By extending the statistical decomposition of Meng (2018), we help shed light on how measurement error does not act to simply alter the magnitude of error but may in fact change the sign as well.  This decomposition motivates the new inverse-probability and doubly robust methodologies in the revised manuscript.
\vspace{5mm}

\item {\it I noticed quite a few rough edges.}

\begin{itemize}
\item {\it Page 4, line 4: The summation variable should be j.}

\vspace{5mm}
We thank the reviewer for pointing this out.  This has been fixed.
\vspace{5mm}

\item {\it Page 8, middle: It should not be difficult to guess what "SIR" means (if one doesn't know), but the author should explain it.}

\vspace{5mm}
We thank the reviewer for pointing this out.  We now mathematically define the susceptible-infected-recovered (SIR) model in Section 3.2 as well as the susceptible-exposed-infected-recovered (SEIR) in Section 4.2.  The SEIR model motivates a new compartmental model that uses daily demographic strata-level COVID-19 death data to estimate daily COVID-19 case counts.  These models are then used in a novel doubly robust method to estimate weekly active infection rates over time.
\vspace{5mm}

\item {\it Figure 3: The caption should explain the dashed curves in part B.}

\vspace{5mm}
We thank the reviewer for pointing this out.  These are the ratio estimator and effective reproductive rate estimator under different relative sampling fractions $M = f_1/f_0$. We see the bias increases as a function of the relative sampling fraction.  We now explain the dashed curves in Figure 3a and 3b.
\vspace{5mm}

\item {\it Page 14, bottom: I wonder about the validity of the estimates of specificity and sensitivity from the Santa Clara study. I recall seeing criticism when the results appeared, but I may be mistaken.}

\vspace{5mm}
We agree that the study has come under significant criticism.  We have updated our choices of specificity and sensitivity to mirror the recent literature on specificity and sensitivity of RT-PCR testing which has grown substantially over the past few months.
\vspace{5mm}
\end{itemize}

\end{enumerate}
\newpage

{\bf Response to AE3}
\begin{itemize}
\item {\it I took some time to read the paper this evening.  The problem considered in this paper is very important and I appreciated that someone is looking into it.  However, I am not sure AOAS is the best home for the paper.  In my humble opinion, epidemiology or general science journals, or maybe medical journals, might be a better fit.  It might receive even more attention if published in those journals than a statistical journal.}

### 2022-02-09 write-up/letters/cover letter-covid.tex
amanda is the best

Kindly find enclosed the Report titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic,'' by Walter Dempsey.

The report contains novel empirical findings and an investigation of statistical paradoxes arising from the analysis of reported coronavirus case-counts that are being used by governments to guide policies and decision making. We have discovered an estimator and statistical decomposition of the associated error that shows the 21.5 million tests conducted in the US provides information on COVID-19 prevalence equivalent to a simple random sample of size 92. Analogous decompositions for the rate of change in prevalence and the effective reproduction number are used to demonstrate how the magnitude and sign of estimation bias varies over time, a bias which has strong policy implications. Overall, our findings demonstrate the statistical limitations of increased testing in the presence of selection bias and measurement error as well as how easy it is for researchers analysing reported coronavirus case-counts to not only fool themselves but be confident in their incorrect conclusions. We believe the work presents an important advance in the statistical understanding of the COVID-19 pandemic.

>>>>

Kindly find enclosed the revised version of the manuscript number AOAS2006-048 titled ``Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic.''  We have revised the article to address the referees’ requests about content, exposition, and other points of clarification.  The reports called for a major revision to re-structure the manuscript around a timely data analysis and related methodology, a more careful discussion of COVID-19 molecular versus serological testing, and to more carefully distinguish the current work from past work on the issues of selection bias and measurement error. The new version complies with this request.

The revision focuses on taking the statistical decomposition previously derived and answering the question of what is to be done given potential selection bias and known measurement-error.  To start, we worked directly with the state of Indiana and are one of only five data requests to be approved by the Governor's office.  The newly public dataset reports daily COVID-19 testing, case counts, and deaths broken out by demographic strata.  To address selection bias and estimate active infection rates, we introduce a novel inverse-probability weighting method that uses auxiliary information to form selection propensities.  The necessary auxiliary information comes from Facebook's large-scale symptom survey which provides relevant information on time-varying covariates such as the fraction of individuals in the population reporting COVID-19 symptoms and/or recent contact with a COVID-19 positive individual.  We extend to a doubly robust method that allows researchers to incorporate compartmental models into their estimation.  The statistical error decomposition is extended to each setting and guides our discussion. The revision now goes well beyond a reframing of the classical problems of selection bias and measurement error by providing health policy experts with new tools to leverage auxiliary information in addressing the ongoing COVID-19 pandemic and future infectious disease outbreaks.

We believe the work presents an important advance in how researchers should think about addressing time-varying selection bias to better understanding the COVID-19 pandemic trajectory. Attached please find a detailed response to the referees outlining how the changes to the manuscript address the concerns raised in their reports.
